---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First we load the data, output to be stored in a variable "rdata"
Then we transform the "day" column into a Date variable

```{r,echo=TRUE}
dset <- read.csv("D:/RWSPACE/Project5_A1/activity.csv",header = TRUE)
dset <- transform(dset, date = as.Date(date))
```


## What is mean total number of steps taken per day?

Make a histogram of the total number of steps taken each day
Used the aggregate function to calculate the total number of steps taken each day
hist function for plotting the histogram

```{r,echo=TRUE}
stepsum <- data.frame(aggregate(dset$steps,by=list(dset$date),sum,na.rm=TRUE))
names(stepsum) <- c("day","total.steps")
hist(stepsum$total.steps,xlab = "total number of steps each day",main = "Histogram of total number of steps each day", col="orange")
```

2. Calculate and report the mean and median total number of steps taken per day

```{r,echo=TRUE}
step.mean <- data.frame(aggregate(dset$steps,by=list(dset$date),mean,na.rm=TRUE))
names(step.mean) <- c("day","mean.steps")
step.median <- data.frame(aggregate(dset$steps,by=list(dset$date),median,na.rm=TRUE))
names(step.median) <- c("day","median.steps")
paste("Mean of total number of steps taken per day")
step.mean
paste("Median of total number of steps taken per day")
step.median
```


## What is the average daily activity pattern?

1.A time series plot of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across all days (y-axis)

```{r,echo=TRUE}
interval <- aggregate(dset$steps,by=list(dset$interval),mean,na.rm=TRUE)
names(interval) <- c("interv","average")
with(interval,plot(interv,average,type="l",xlab ="5-min interval",ylab = "Average number of steps",main = "Average number of steps per 5-min interval"))
```

The 5-minute interval, on average across all the days containing the maximum number of steps

```{r,echo=TRUE}
interval$interv[which.max(interval$average)]
```


## Imputing missing values
1.Calculate and report the total number of missing values in the dataset

```{r,echo=TRUE}
total.NA <- length(dset$steps[is.na(dset$steps)==TRUE])
percent.NA <- total.NA/length(dset$steps)
sprintf("NA values represent %1.2f%% of the observations total number",100*percent.NA)
```

2.Imputing the missing values by replace NAs by the mean of that 5-min interval

```{r,echo=TRUE}
imputed <- dset
##Code for imputing missing values
for(i in 1:17568){
      if (is.na(imputed$steps[i])==TRUE){
            indic <- imputed$interval[i]
            imputed$steps[i] <- interval$average[match(indic,interval$interv)]
      }
}
```

3.Make a histogram of the total number of steps taken each day (after imputing NAs)

```{r,echo=TRUE}
stepsum1 <- data.frame(aggregate(imputed$steps,by=list(imputed$date),sum,na.rm=TRUE))
names(stepsum1) <- c("day","total.steps")
hist(stepsum$total.steps,xlab = "total number of steps each day",main = "Histogram of total number of steps each day after imputing NAs", col="green")
```

4.Calculate and report the mean and median total number of steps taken per day

```{r,echo=TRUE}
step.mean_ <- data.frame(aggregate(imputed$steps,by=list(imputed$date),mean,na.rm=TRUE))
names(step.mean_) <- c("day","mean.steps")
step.median_ <- data.frame(aggregate(imputed$steps,by=list(imputed$date),median,na.rm=TRUE))
names(step.median_) <- c("day","median.steps")
paste("Mean of total number of steps taken per day")
step.mean_
paste("Median of total number of steps taken per day")
step.median_
```

We can notice Only NA values have changed into valid values, for the others, no changes were recorded 


## Are there differences in activity patterns between weekdays and weekends?

First, we create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day

```{r,echo=TRUE}
library(dplyr)
weekday <- mutate(imputed,category=weekdays(imputed$date))
for(i in 1:17568){
      if (weekday$category[i]=="samedi"|| weekday$category[i]=="dimanche"){
            weekday$category[i] <- paste("weekend")
      }
      else weekday$category[i] <- paste("weekday")
}
weekday <- transform(weekday,category=as.factor(category))
benchmark <- aggregate(weekday$steps,by=list(weekday$interval,weekday$category),mean)
names(benchmark) <- c("inter","category","avg")
```

Panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r,echo=TRUE}
library(lattice)
xyplot(avg ~ inter | category, data = benchmark,layout=c(1,2), type="l",lty=1)
```

We notice that the maximum average number of steps isreached in weekdays, but in general, and accross the day, the activity is more important in weekends than in weekdays 